\chapter{Introduction}\label{ch:introduction}
\epigraph{
- Homer, is it the way you pictured PhD life?\\
- Yeah, pretty much, except we drove in a van solving mysteries.
}{Could be Homer Simpson}

\begin{addmargin}[5em]{5em}
This thesis links the research fields of data mining, constraint learning
and logic programming. We introduce each of them in turn and provide an overview of the
contributions of the thesis and
its general structure.
\end{addmargin}

\section{Data mining}
The ability to write down a model and to simply push a button to get the
answer is an appealing idea. It is one of the holy grails of
Artificial Intelligence, called Declarative Programming, where one
specifies what the problem is and not how to solve it. Today in the age of data, we would like
to state what type of knowledge we are looking for, feed the data in and, as a result,
(almost magically) find new insights into the data. What we do not really want is to write
a lot of tedious code. Unsurprisingly, this is a hard challenge. 


The \textit{knowledge} extracted from the data can manifest itself in
multiple forms. For example, an interesting piece of data,
\textit{typically a substructure}, is referred to as a \textit{pattern} \parencite{han_book}. 
The most known pattern learning task is 
\textit{frequent pattern mining}, or pattern mining for short \parencite{survey_han}. 
At a high level it can be formalized as follows: 
let $D$ be a dataset, $P$ be the language of allowed patterns and
$\phi$ be a predicate of
interest, the pattern mining problem is the problem of finding
all patterns $p$ in $P$ such that $\phi(p,D)$ holds.
Simply speaking, we enumerate objects that have a certain property of
interest.
Various measures of relevance or ``interestingness'' of a pattern exist \parencite{tias_topk}.


Many Data Mining problems, and especially pattern mining
problems, are \textbf{combinatorial search problems}, which in turn have two main components, namely problem modelling and search: 

\begin{center}
  Combinatorial problem = Model + Search
\end{center}

Therefore, it seems natural to model pattern mining problems 
using systems designed to perform search in declarative manner, such
as \acrlong{cp} \parencite{handbookcp} or \acrlong{asp}
\parencite{whatisasp}. %(which we discuss later in detail).

\section{Constraint programming} 
\acrlong{cp} is a declarative methodology for solving combinatorial
problems. The essence of this methodology consists of writing a set of
constraints, such as $X + Y > Z$, and inferring a solution to them, such
as $X = 1, Y = 2$ and $Z = 0$. This set of constraints is referred as
a \textit{specification} or a \textit{model}; for this, multiple \acrlong{cp}
languages exist, such as MiniZinc \parencite{minizinc} or Essence
\parencite{essence}. A \textit{solution} for
a model is an assignment, as for example, the one we just indicated, that satisfies
\textit{all} constraints. This general methodology is declarative,
since a user writes a set of constraints corresponding to the
properties of her problem and a \acrshort{cp} solver is looking for a
solution: this clearly demonstrates that it is a task of the solver
to figure out \textit{how} to find a solution, while it is the user's
responsibility to specify \textit{what} the problem is in terms of
writing these constraints.

\section{Constraint learning}
Constraint Learning is the research field concerned with finding
constraints that hold in data \parencite{constraint_learning,QUACQ,Conacq}.
%Let us outline in simple terms, what constraint learning is about.
Similar to the pattern matching problem, we can define 
the constraint learning problem as follows:
%Explaining what constraint learning in term of the previous section:
let $D$ be a dataset, $C$ be a set of interesting properties (constraints), and
$\phi$ be a satisfaction function, 
then the problem is to find all properties (constraints) $c$ in $C$ such that $\phi(D,c)$ is satisfied.
\begin{example}[Sudoku constraint learning]
    A well-known Sudoku problem might be hard to model for a person
    new to constraint programming. A possibility how to obtain a model
    is to learn from examples. This can be done by providing:
    \begin{itemize}
        \item A set of Sudoku solutions
        \item A set of invalid Sudoku combinations
        \item An expressive constraint language 
    \end{itemize}
Then, a property of interest is a constraint that holds for all
    positive Sudoku combinations (all solutions) and for none of
    the invalid combinations. For example, this can be a constraint of
    the form
\begin{equation*}
    \forall i: \sum_j x_{ij} = 45
\end{equation*}
    Which shows that the sum of elements in each column is equal to 45
    (simply the sum of numbers from 1 to 9).
\end{example}

In case of pattern mining, we enumerate objects on which the property
of interest holds: we find graphs or sequences that occur often enough
in the data and have certain desired topological properties (for example, being a
cyclic graph or have ascending characters). And in constraint learning we find interesting
properties that hold in the data (or its parts): given a set of
characters, for example, we discovered that the "ascending constraint" holds, i.e.,
characters are sorted or we found that givens graphs are all cyclic.
Therefore, we consider constraint learning and pattern mining to be interconnected, and both problems play a central role in this thesis.

\section{Logic programming}
In his seminal work Robert \textcite{kowalski} proclaimed:
\begin{center}
  Algorithm = Logic + Control.
\end{center}

%What he meant is quite simple, 
The intuition behind this statement is that
you can turn logical rules into a computational process. 
Why is this important for us? Logic deals with
relations and relational structures, it is naturally designed to
work with relational data. Therefore, if we can combine logic
with computation, we obtain a computational engine that can work in
our relational learning setting.

\begin{example}[Modeling Sushi preferences with logic]
There are three people: Ronald, Peter and
Alina. We know that people who are not allergic to fish and not vegans like sushi; 
furthermore, we know that Peter is allergic to fish and Ronald is
    vegan. 
Who likes sushi then? We can model this puzzle as a
simple logical implication:

\begin{equation*}
    \begin{aligned}
& \textit{likes\_sushi}(X)~{:}{-}~\textit{person}(X),~\text{not}~
        \textit{allergic}(X), ~\text{not}~\textit{vegan(X)}. \\
&       \textit{allergic(peter)}. \\
&       \textit{vegan(ronald)}.
    \end{aligned}
\end{equation*}

The following notation in Prolog (Programming in logic)
\parencite{prolog_original} mimics the logical implication:

\begin{equation}\label{eq:sushi}
  \forall X: \textit{person}(X) \wedge \text{not }
    \textit{allergic}(X) \wedge \text{not } \textit{vegan}(X)
  \implies \textit{likes\_sushi}(X)
\end{equation}
\end{example}
Therefore, we can think of Logic Programming as an executable Logic, that
we can use to model and solve relational problems.

In this thesis we use modern Logic Programming engines, such
as \acrlong{asp} (\acrshort{asp}) \parencite{ASPbook,whatisasp} and
\acrlong{idp} (\acrshort{idp})
\parencite{idp} %(we explain them in details in the following chapter)
that naturally support constraints and various types of logical
inference.

\section{Relational and logical learning}
Typical machine learning and data mining systems work with
propositional representations and logical and relational learning has
arisen to overcome this limitation by focusing on expressive knowledge representation languages such as first order logic \parencite{luc_book}.

Classical machine learning algorithms such as induction of decision trees
\parencite{decision_trees} or Bayesian networks \parencite{pearl} work
with limited representations (which correspond to a
propositional language). This representation does not allow to
represent naturally problems involving multiple complex objects
together with the relationships that hold between them. This limitation has been
pointed out by a number of researchers, e.g., in the seminal work of
\textcite{plotkin}, who has introduced  the use of first order logic
for machine learning. This approach allows to
represent a non-fixed amount of complex objects together with their
relationships -- we refer to this problem setting as \textit{relational}.

Let us indicate why it is important to look into the relational
machine learning setting. To do so, we will describe a problem where
the number of entities and relationships between them is not fixed,
which makes it complicated if not practically impossible to employ the
standard machine learning techniques based on  fixed feature vectors
or attribute-value representations, while using logical and relational models appears natural and elegant. Further in the chapters we elaborate on other problems and examples.

A very well familiar to a reader, a schedule is an example
of relational data. 
\begin{example}[Examination room scheduling]
To make this example concrete, assume that students need to be assigned to
auditoria for examinations with professors. Typically, making such a
schedule involves dealing with interconnected constraints:
    \begin{itemize} 
    \item certain exams
must have enough gap time between them 
    \item rooms can be shared by at most
    a few courses 
\item professors might have their own personal preferences
\item university administration might enforce certain administrative and 
legal constraints 
    \end{itemize} 
    
    A possible approach is to learn these 
constraint by recovering them from
a set of positive and negative examples. The number of constraints is
not set in advance and the constraint language is typically a rich
relational language, for example, a constraint saying that no room is
shared by more than three courses at the same time can be expressed in a form of the first
order language as
\begin{equation*}
    \forall X,T: \textit{room(X)} \wedge \textit{time(T)} \implies \{ C : \textit{examination}(X,C,T) \wedge \textit{course(C)} \} \leq 3.
\end{equation*}
\end{example}

%   A very well familiar to a reader spreadsheet with its formulae is an
%   excellent example of relational data: it contains a number of
%   relations or \textit{tables}, and their number is never fixed in
%   advance, since a spreadsheet might contain an arbitrary number of
%   them. Also, it combines both textual and numeric data. Furthermore,
%   formulae and constraints over the tables play a role of relationships
%   we have described earlier.
These properties of interconnectedness and expressivity of the constraints make
problem hard to model using traditional propositional approaches,
while this can be clearly seen as a first order problem. This gives us
confidence that for a number of important problems it is essential to
look into richer knowledge representation languages for machine
learning.

\section{Contributions}
Given the abundance of relational data (such as spreadsheets, graphs,
databases and logical structures) for various machine learning and data
mining problems, and the success in development of declarative solvers (such
as \acrshort{asp}, \acrshort{idp}, \acrshort{cp} and \acrshort{lp}) a
natural question arises: \textit{``How can we declaratively model and
solve relational data mining problems using logic programming
techniques?''}. 

The \textbf{main contribution} of this thesis is in the application of
modern logic programming languages, such as \acrshort{asp} and
\acrshort{idp}, to these novel relational data mining problems.


Concretely, we investigate how the following data mining problems
can be modelled and solved using a declarative approach based on
logic programming.
\begin{itemize}
    \item \cone: \textbf{Relational Data Factorization.} What are the challenges and advantages of generalizing
    classic data mining problems, such as the Boolean Matrix
    Factorization problem, into the relation setting?
  \item \ctwo: \textbf{Tabular Constraint Learning.} How can the constraint learning problem be formulated
   and modelled in the relational setting, where data is
   represented as spreadsheets, i.e., connected relations, and constraints are
   tabular Excel-like formulae?
  \item \cthree: \textbf{ASP Sketching.} How can the problem of ASP
      Sketching, where one marks certain parts of the
        program to be uncertain, be modelled and solved using our
        declarative approach? 
   \item \cfour: \textbf{Structured Frequent Pattern Mining.}
%  Similarly to unstructured constraint-based itemset mining: 
    How can we apply declarative relational mining
    techniques, such as logic programming, to structured frequent pattern mining, such as sequence, graph
    and query mining?
\end{itemize}

Addressing \cone, \textbf{Chapter} \ref{ch:ReDF} presents the novel problem of Relational Data
Factorization. It demonstrates how this problem generalizes various
data mining problems such as Database Tiling and Boolean Matrix
Factorization. We start with the most basic setting and demonstrate
how various problems can be modelled within the framework by adding
or modifying the constraints. Then, we show how the framework can be
used for new and interesting applications and propose Answer Set
Programming as a method to solve the problem in the general case.
For each problem we provide extensive experimental evidence,
including solver parameter tuning. The chapter consists of the
research previously published in the following paper:

\begin{addmargin}[2em]{2em}

Sergey Paramonov,  Matthijs van Leeuwen, Luc De Raedt: Relational data
factorization, Machine Learning, Springer, December 2017, Volume 106,
    Issue 12, pp 1867–1904.

\end{addmargin}



With regards to \ctwo,  \textbf{Chapter} \ref{ch:TaCLe} introduces  the novel problem of
Tabular Constraint Learning and the system, called TaCLe (pronounced
as the word ``tackle''). The problem setting is straightforward to
explain: we have an Excel file with formulae. The file is imported as
CSV. The formulae are lost. Can we reconstruct them? We can already 
see that the problem setting is different from the standard machine
learning problem. In a spreadsheet, columns are no longer variables
and rows are no longer records. Textual and numeric data are mixed.
Spreadsheet functions, such as Fuzzy-lookup, are unorthodox and unseen in
the constraint programming and constraint learning research communities.

The chapter consists of the research previously published in the following papers:

\begin{addmargin}[2em]{2em}
Samuel Kolb (*), Sergey Paramonov (*), Tias Guns, Luc De Raedt:
  Learning constraints in spreadsheets and tabular data. Machine
  Learning 106 (9-10), pages 1441-1468, 2017.


Sergey Paramonov (*), Samuel Kolb (*), Tias Guns, Luc De Raedt:
TaCLe: Learning constraints in tabular data. 
 Proceedings of the 2017 ACM on Conference on Information and
    Knowledge Management, CIKM 2017, pages 2511-2514, Sheridan
    Communications, Conference on Information and Knowledge Management
    (CIKM), Singapore, 6-10 Nov 2017.
\end{addmargin}




%   we look into a more practical setting.
%   Typically, machine learning and data mining algorithms' assumptions make the i.i.d.
%   (independent identically distributed) assumption, which of course
%   breaks in the relational setting. If we look at a spreadsheet, it 
%   typically consists one or more tables with relationships between them or
%   within the table itself, usually
%   in a form of a formula, a macro or of some other constraint (e.g., a
%   C\# or VB script). If we
%   want to detect these connections between tables or their parts, we
%   have to make the model able to work with spreadsheet data, 
%   formulae and tabular constraints. This setting, referred to as the
%   \textit{Tabular constraint learning}, breaks the standard
%   machine learning and data mining algorithms and 
%   makes the problem essentially relational.

Addressing \cthree:  \textbf{Chapter} \ref{ch:sketching} introduces the novel problem of
Sketched Answer Set Programming. The idea of sketching takes root in
imperative programming, when a function in C or Java has a part of it,
typically a constant, left unspecified. A user then provides a number
of examples for this sketched constant to be filled in with a
correct value satisfying the examples. Inspired by this approach, we
introduce Sketched Answer Set Programming, where a user writes a
regular program but with certain parts such as constants, predicates
or operators are left uncertain. Then, we rewrite this sketched
program into a regular \acrlong{asp} program, i.e., we use a standard ASP solver
to complete sketches.


The chapter consists of the research previously published in the following papers:
\begin{addmargin}[2em]{2em}
  Sergey Paramonov, Christian Bessiere, Anton Dries, Luc De Raedt:
  Sketched Answer Set Programming. CoRR abs/1705.07429, 2017.
\end{addmargin}




%   assume we already have a logic programming or a
%   constraint model of a relational
%   data mining problem due to relational dependencies in the data which 
%   force to give up the i.i.d. assumption and, as a consequence,
%   standard techniques. When something changes over time or due to changes in the specification, we would like
%   to be able to indicate uncertainty in some parts of the model and use
%   the system itself. These
%   uncertain or, \textit{open}, parts are called \textit{sketched} and
%   the whole sketched model is referred to as a \textit{sketch} and the related
%   problem as \textit{constraint sketching}. We propose the problem of and
%   a learning technique for \acrlong{skasp}.


Finally, with respect to \cfour, \textbf{Chapter} \ref{ch:StructuredMining} introduces a logic
based approach to structured pattern mining. The idea to apply general
solvers to pattern mining takes root in the work of
\textcite{declrativeapproach}, where Constraint Programming is applied to
Itemset Mining. Contrary to itemset mining, where objects do not have
any particular structures, structured mining studies mining patterns
in sequences, trees, graphs, etc. First, we introduce a general
mathematical model of general frequent pattern mining based on the logical formulation of
the constraints. Second, we show that the model can be hybridized by
using our generic framework together with specialized solvers
developed for particular pattern mining problems.

The chapter consists of the research previously published in the following papers:
\begin{addmargin}[2em]{2em}
Sergey Paramonov, Matthijs van Leeuwen, Marc Denecker, Luc De Raedt:
An Exercise in Declarative Modeling for Relational Query Mining.
Lecture Notes in Computer Science, volume 9575, pages 166-182,
    Springer, Inductive Logic Programming, Kyoto, 20-22 August 2015.

Tias Guns, Sergey Paramonov, Benjamin Negrevergne: On declarative
    modeling of structured pattern mining, AAAI Workshop on
    Declarative Learning Based Programming, Phoenix, Arizona USA, 13
    February 2016.

Matthias van der Hallen, Sergey Paramonov, Michael Leuschel, Gerda
    Janssens: Knowledge representation analysis of graph mining,
    Bogaerts, Bart; Harrison, Amelia (eds.), Proceedings of ASPOCP
    2016, pages 55-76, Workshop on Answer Set Programming and Other
    Computing Paradigms, New York, 16 October 2016.

Sergey Paramonov, Chen Tao, Tias Guns: Generic mining of condensed
    pattern representations under constraints, Young Scientist's
    Second International Workshop on Trends in Information Processing
    (YSIP2), pages 168-177, CEUR Workshop Proceedings, Young
    Scientist's International Workshop on Trends in Information
    Processing, Dombai, Russia, 16-20 May 2017.

Sergey Paramonov, Daria Stepanova, Pauli Miettinen:
Hybrid ASP-Based Approach to Pattern Mining.  
Rules and Reasoning - International Joint Conference, RuleML+RR 2017,
    Springer Lecture Notes in Computer Science (LNCS), pages 199-214,
    Springer, 2017, RuleML+RR 2017: International Joint Conference on
    Rules and Reasoning, July 12-15, London, UK.
\end{addmargin}




%   if we look at the area of itemset mining, where \acrlong{cp}
%   has been successfully applied, then the natural question to ask is:
%   if \acrshort{cp} has been successful for unstructured, or
%   non-relational, pattern mining, can we model its \textit{structured},
%   or \textit{relational} version, using logic programming? To answer
%   this question we look into \textit{sequence}, \textit{graph} and
%   \textit{query mining} and attack them using logic programming and
%   hybrid approaches based on logic programming.

%   \paragraph{Structure of the thesis}
%   \textbf{Chapter} \ref{ch:background} introduces background information
%   on Logic, Answer Set and Programming and Pattern Mining.

\textbf{Chapter} \ref{ch:conclusions} summarizes the thesis and
discusses future work possibilities.

\section{Datasets, code and experimental results}
In order to make results repeatable and to follow the Open Source
spirit, we have opened and published the datasets, use-cases,
meta-data and other useful information.
They can be found in the following GitHub repositories:
\begin{itemize}
\item \url{https://github.com/SergeyParamonov/TaCLe}
\item \url{https://github.com/SergeyParamonov/sketching}
\item \url{https://github.com/SergeyParamonov/LGM}
\end{itemize}
\cleardoublepage
% vim: tw=70 nocindent expandtab foldmethod=marker foldmarker={{{}{,}{}}}
