\chapter{Relational data factorization}\label{ch:ReDF}
\newcommand{\redfpath}{chapters/ReDF/tex_files}


Motivated by an analogy with matrix factorization, we introduce the problem of factorizing relational data. In matrix factorization, one is given a matrix and has to factorize it as a product of other matrices. In \acrlong{redf} (\acrshort{redf}), the task is to factorize a given relation as a conjunctive query over other relations, i.e., as a combination of natural join operations. Given a conjunctive query and the input relation, the problem is to compute the \emph{extensions} of the relations used in the query. Thus, relational data factorization is a relational analog of matrix factorization; it is also a form {\em inverse} querying as one has to compute the relations in the query from the result of the query.  The result of relational data factorization is neither necessarily unique nor required to be a lossless decomposition of the original relation. Therefore, constraints can be imposed on the desired factorization and a scoring function is used to determine its quality (often similarity to the original data). Relational data factorization is thus a constraint satisfaction and optimization problem. We show how answer set programming can be used for solving relational data factorization problems. 


\section{Introduction}
\input{\redfpath/introduction} 
\section{Relational data factorization}
\label{section:framework}
\input{\redfpath/formalization}

\section{Application to data mining problems}
\label{section:dm_problems}

In this section we show that the ReDF framework generalizes a wide range of data mining tasks and provides a truly declarative modeling approach for relational data factorization. We introduce a range of constraints and optimization criteria that can be used in practice. The data mining tasks studied include tiling \parencite{tiling},  Boolean Matrix Factorization (BMF) \parencite{dbp}, discriminative pattern mining \parencite{DBLP:conf/pkdd/KnobbeH06}, and block-diagonal matrix forms \parencite{blockdiagonal}.%and block-models \parencite{blockmodelFirst,blockmodelsWhite,blockmodelKemp,conf/aaai/KempTGYU06}.%  are discussed and evaluated in Appendix \ref{section:other_dm_tasks}).

\subsection{Tiling}
\label{subsection:tiling}
\input{\redfpath/DMproblems}
\section{Beyond classic problems}
\label{subsection:beoynd}
\input{\redfpath/beyond_classic_problems}

\section{Implementation}
\label{section:implementation}
\input{\redfpath/implementation}

\section{Experiments}
\label{section:experiments}
\input{\redfpath/experiments}

\section{Evaluating solver performance}  \label{subsec:evaluatingsolver}
\input{\redfpath/solver_tuning}

\section{Visualizations for an application of purely relational ReDF}\label{appendix:application_purel_relational}
\input{\redfpath/application_purely_relational_factorization_visualization.tex}
\section{Maximum $k$ set coverage analysis of greedy strategy}\label{appendix:k_set_coverage_analysis}
\input{\redfpath/greedy_k_coverage_analysis.tex}

\section{Related work}
\label{section:related_work}
\input{\redfpath/RelatedWork}
\section{Conclusions}
\label{section:discussion}
\input{\redfpath/discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Keep the following \cleardoublepage at the end of this file, 
% otherwise \includeonly includes empty pages.
\cleardoublepage

